---
title: "PSY 4802- Quiz 4"
author: "Amanda Mae Woodward"
date: "2024-11-19"
output: html_document
---
```{r}
library(ggplot2)
```

For this assignment, we're going to use data that were collected by Rentfrow et al on the big five personality traits across regions of the U.S.

1) import the ANOVA OCEAN by state dataset 
```{r}
oceanData <-  read.csv("~/Desktop/FA24 Syllabi:Schedules/PSY4802_Using_R_to_Create_Reproducible_Research/Quiz/ANOVA OCEAN by state.csv")
```

2. Find the correlation between openness scores (o) and extraversion (e)
```{r}
cor.test(oceanData$E, oceanData$O)
#### the correlation is both relatively weak and negative, -0.16. 
```

3. Create a graph to accompany your analysis. 
```{r}
# decorate graph 
ggplot(oceanData, aes(x = E, y = O)) +
  geom_point(color = "#F34213", size = 3, alpha = 0.7) +  
  geom_smooth(method = "lm", se = TRUE, color = "#08A045",size = 0.7) +
  labs(
    title = "Relationship of Extraversion and Openness",
    x = "Extraversion ",
    y = "Openness"
  ) +  
  theme_minimal() +  
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),  
    panel.grid.major = element_line(color = "gray80", linetype = "dotted"),  
    panel.grid.minor = element_blank()
  )  
```

4. Create a subset of your data containing only east coast and midwest data.
```{r}
regionalData <- subset(oceanData, Region %in% c("NE", "MW"))
```

5. Using this dataset, run a t test to determine if the agreeableness (A) levels differ between the midwest and east coast
```{r}
midwestDat <- subset(regionalData, Region == "MW")

eastCoastDat <- subset(regionalData, Region == "NE")

regionalT <- t.test(midwestDat$A, eastCoastDat$A, var.equal = FALSE)
print(regionalT)
```
Based on the results of our t-test, we can state that the mean difference in agreeableness between residents of the Midwest and East Coast is not equal to zero and is in fact significantly different. 

6. create a graph to illustrate your findings.
```{r}
ggplot(regionalData, aes(x = Region, y = A, fill = Region)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.4, fill = "#F7B32B") +
  labs(
    title = "Distribution of Agreeableness Levels by Region",
    x = "Region",
    y = "Agreeableness"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("NE" = "#17BEBB", "MW" = "#D62246")) +
  theme(legend.position = "none",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5) 
        )
```

7. create a subset of the original dataset containing data from the west and south. 
```{r}
SW_subset <- subset(oceanData, Region %in% c("S","W"))
```

8. run a correlation between region and openness. 
```{r}
corrResult <- aov(O ~ Region, data = oceanData)
summary(corrResult)
```

9. create a graph to accompany this analysis
```{r}
ggplot(oceanData, aes(x = Region, y = O, fill = Region)) +
  geom_boxplot(outlier.color = "#3D0B37") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 5, fill = "white", color = "#1F363D") +
  labs(
    title = "Openness Across Region",
    x = "Region",
    y = "Openness"
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none")
```

10. Going back to your original dataset, run an ANOVA to determine if neuroticism (N) differs across region. create a graph to accompany your analysis. 
```{r}
# Run ANOVA
anova_result <- aov(N ~ Region, data = oceanData)
summary(anova_result)
```
Our ANOVA result is significant, and mean reported neuroticism does differ significantly across regions. 
```{r}
ggplot(oceanData, aes(x = Region, y = N, fill = Region)) +
  geom_boxplot(outlier.color = "#DE3C4B", alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", color = "black", fill = "white") +
  labs(
    title = "Neuroticism Across Region",
    x = "Region",
    y = "Neuroticism"
  ) +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12), 
   legend.position = "none") # suppress since already labeled on x 
```