---
title: "Quiz 2"
author: "Amanda Mae Woodward"
date: "2024-10-15"
output: html_document
---

This quiz will focus on data frame manipulation, however, you may need to rely on skills you've learned in prior weeks. 

1. load the starwars dataset from tidyverse
```{r}
library(tidyverse)
data(starwars)
```

2. separate the films column so that there is one film in each column (don't worry about the "c(" and ) at the end for now)
```{r}
divideByFilm <- starwars %>% 
  separate(films, into = paste0("film", 1:10), sep = ",", fill = "right")
```

3. alter the data from wide data to long data
```{r}
longFilm <- divideByFilm %>% pivot_longer(cols = starts_with("film"),
                                          names_to = 'film_number',
                                          values_to = 'film',
                                          values_drop_na = 1)
```

4. in the column where you stored the films, rename the levels so each film name is consistent (e.g. get rid of the "c(" and ")" )
```{r}
filmsLonger <- longFilm %>% 
  mutate(film = str_replace_all(film, "c\\(|\\)", ''))
```
I Googled for help with this question a lot more than usual. Now I know that R uses backslash as an escape character! 

5. Create a column in your dataset that indicates whether a character was in the first starwars film, *Star Wars Episode IV: A New Hope* 
```{r}
filmsLonger <- filmsLonger %>% 
  mutate(inFirstFilm = if_else(str_detect(film, "Star Wars: Episode IV: A New Hope"), TRUE, FALSE))
```

6. BMI is calculated as mass in Kg/height in meters. You currently have mass in Kg and height in cm. Create a column that contains each character's BMI. 
```{r}
filmsLonger <- filmsLonger %>% 
  mutate(height_m = height / 100, 
         bodyMassIndex = mass / (height_m ^2))
```

7. create a dataframe that contains only the name, hair color, and eye color for each character. 
```{r}
charInfo <- starwars %>% 
  select(name, hair_color, eye_color)
```

