---
title: "Quiz 3"
author: "Amanda Mae Woodward"
date: "2022-10-30"
output: html_document
---

```{r}
# load necessary libraries 
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(viridis)
```

1. Load the spotify dataset (available in github repository).
```{r}
# github permalink won't work? 
spotifyDat <- read.csv("~/Desktop/FA24 Syllabi:Schedules/PSY4802_Using_R_to_Create_Reproducible_Research/Quiz/spotifyDat.csv")
```

2. Create a bar graph of the artist column (Hint: use theme(axis.text.x = element_text(angle = 90)) to rotate the x axis).
```{r}
ggplot(spotifyDat, aes(x = artist)) +
  geom_bar() + 
  theme(
    axis.text.x = element_text(angle = 90)
    )
```

3. Add genre as a third variable in your graph.
```{r}
ggplot(spotifyDat, aes(x = artist, fill = genre)) + 
  geom_bar() + 
  theme(
    axis.text.x = element_text(angle = 90)
    )
```

4. Use color brewer to add color to your graph in 3. 
```{r}
ggplot(spotifyDat, aes(x = artist, fill = genre)) + 
  geom_bar() + 
  scale_fill_brewer(palette = "Set2") + 
  theme(
    axis.text.x = element_text(angle = 90)
    )
```

5. Place the legend for the graph you made in 3 on the bottom of the graph. 
```{r}
ggplot(spotifyDat, aes(x = artist, fill = genre)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set3") +
  theme(
    axis.text.x = element_text(angle = 90), 
    legend.position = "bottom"
    )
```

As you probably recognized, the graph you made is really hard to read.

6. Separate the genre column into multiple columns (e.g. so each genre listed is in a unique column)
```{r}
spotifyDat <- spotifyDat %>% 
  separate(genre, into = paste0("genre", 1:3), sep = ",", fill = "right")
```

7. Use these genre columns to create a dataset in long format. 
```{r}
longSpot <- spotifyDat %>% 
  pivot_longer(cols = starts_with("genre"), names_to = "genre_type", values_to = "genre") %>%
  filter(!is.na(genre)) # ok to remove NaN's? 
```

8.use facet wrap to make graphs of artist by genre 
```{r}
ggplot(longSpot, aes(x = artist, fill = genre)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ genre)
```

9. Create a graph that includes the variables tempo and dancability. 
```{r}
ggplot(longSpot, aes(x = tempo, y = danceability)) +
  geom_point()
```

10. Add popularity as a third variable. 
```{r}
ggplot(longSpot, aes(x = tempo, y = danceability, color = popularity)) +
  geom_point()
```

11. Use viridis as a color in your graph from #10
```{r}
ggplot(spotifyDat, aes(x = tempo, y = danceability, color = popularity)) +
  geom_point() +
  scale_color_viridis(alpha = 0.6, option = "H")
```

