---
title: "Dataset manipulation pt 2"
author: "Amanda Mae Woodward"
date: "2024-10-15"
output: html_document
---

We're going to use this part of class to review how to do the learning outcomes from the dataset manipulation notes. 

### load data
1. Open the palmerpenguins library
```{r}
library(palmerpenguins)
```

2. Open the penguins dataset
```{r}
data(penguins)
```

3. Explore the raw dataset. What can you find out about it? 



4. Are the data in long or wide format? 
```{r}
summary(as.factor(penguins_raw$`Sample Number`))
summary(as.factor(penguins_raw$Species))
summary(as.factor(penguins_raw$`Individual ID`)) # this is what we will use 
# data is in long format
```
long data, but missing measurements for some penguins 


5. let's practice going from one format to the other. 

we are going from long to wide data. 
using flipper length as the variable we want to measure 

```{r}
library(tidyverse) # pipeline
widePengy <- penguins_raw %>%
  pivot_wider(id_cols = 'Individual ID ', names_from = "Sample Number", values_from = "Flipper Length (mm)")

tinyPenguin <- penguins_raw %>% select("Individual ID", "Sample Nnumber", "Flipper Length (mm)")

tinyPenguinsWide <-  tinyPenguin %>% pivot_wider(id_cols = "Individual ID", names_from = "Sample Number", values_from = "Flipper Length (mm)")
```

6. create two datsets. One should have the individual id, region, and island. The other should have individual id, species, flipper length, culmen length, culmen depth, and body mass. 
```{r}
penguinSubOne <- select(penguins_raw, c("Individual ID", "Region","Island"))
  
penguinSubTwo <- select(penguins_raw, c("Individual ID","Species","Flipper Length (mm)","Culmen Length (mm)","Culmen Depth (mm)","Body Mass (g)"))
```

7. Join the two datasets you made together.
```{r}
a <- left_join(penguinSubOne, penguinSubTwo)
# warning is telling you some ID's have multiple pieces of information 
```

8. Separate the species column into the common name and the scientific name 
```{r}
separate(penguins_raw, col = 'Species', into = c("commonName","scientificName"), (sep = " "))
```

9. change the levels of clutch completion to indicate whether the penguin has two eggs or does not have two eggs
```{r}

```

10. create a column that has body mass in kilograms.
```{r}

```

11. copy island to a new column. Recode this variable so that each island is renamed to be just the first initial
```{r}

```

