---
title: "Dataset Manipulation Assignment"
author: "Amanda Mae Woodward"
date: "2024-10-08"
output: html_document
---
```{r load packages}
library(tidyverse)
```

```{r}
# Load necessary libraries
library(ggplot2)

# Generate random data
set.seed(123)
data <- data.frame(values = rnorm(1000, mean = 50, sd = 10))

# Create a poorly designed histogram
bad_histogram <- ggplot(data, aes(x = values)) +
  geom_histogram(bins = 50, fill = "green", color = "purple") +  # Too many bins
  labs(title = "Histogram of Values",  # Misleading title
       x = "Value",  # Vague x-axis label
       y = "Number") +  # Misleading y-axis label
  theme_minimal() +  # A clean theme that doesn't fit the context
  xlim(0, 100) +  # Misleading x-axis limits
  geom_vline(aes(xintercept = 70), color = "red", linetype = "dashed", size = 1) +  # Arbitrary line
  geom_text(aes(x = 70, label = "Mean?"), y = 30, color = "black", size = 5)  # Confusing text label

# Display the histogram
print(bad_histogram)

```

#Part 1
1.Open the sleep dataset in R 
```{r}
data(sleep)
```

2. Is it long or wide format? How do you know? 

  This data is in long format, as subjID's and group repeat in rows 
  
3. Transform this data so that it is in the other format. 

```{r}
wideSleep <- sleep %>%
  pivot_wider(names_from = group, 
              values_from = extra)
```

4. Transform your data back to its original format. 

```{r}
longSleep <- wideSleep %>%
  pivot_longer(cols = -ID, 
               names_to = "group", 
               values_to = "extra",
               names_prefix = "group_")
```

5. Open the Starwars dataset
```{r}
data(starwars)
```

6. Create homeworld as a factor. How many levels does it have? 
```{r}
as.factor(starwars$homeworld)
```

7. Reorder homeworld so that Naboo is the reference group. 
```{r}
homeworld <- as.factor(starwars$homeworld)
homeworld <- relevel(homeworld, ref = "Naboo") 
levels(homeworld)
# googled and found relevel() function/syntax on stackoverflow 
```

8. Use mutate to create a column transforming kg to pounds. (hint: you may need to look up the conversion)
```{r}
starwars <- starwars %>% 
  mutate(massLbs = mass * 2.20462)
```

9. Print a summary of your column from #8
```{r}
summary(starwars$massLbs)
```

#Part 2
10. open the billboard dataset.
```{r}
data(billboard)
```

11. are the data in this dataframe in long or wide format? explain how you know. 

This data is in wide format because there is one single song per row, and the weeks span many columns 

12. change the data from its original format to the other version. 
```{r}
bill_long <- billboard %>% 
  pivot_longer(cols = starts_with("wk"), 
               names_to = "week", 
               values_to = "rank")
```

13. change it from your altered version in #3 back to its original state. 
```{r}
bill_wide <- bill_long %>% 
  pivot_wider(names_from = "week", 
              values_from = "rank")
```

14. Separate the date entered column into three columns: one for year, one for month, and one for day. 
```{r}
billboard <- billboard %>% 
  separate(date.entered, 
           into = c("year", "month", "day"), 
           sep = "-")
```
This last question took some googling and scraping through the tidyverse documentation... 
