---
title: "ggplot part 1 assignment"
author: "Amanda Mae Woodward"
date: "2024-10-15"
output: html_document
---
1. Open the starwars dataset from tidyverse
```{r}
library(tidyverse)
data(starwars)
```

2.Create a graph of character's eye color. (Hint: pay attention to the class of data)
```{r}
ggplot(starwars, aes(eye_color)) + 
  geom_bar()
```

> Credit to https://coolors.co/ for the fun color palettes 

3. Customize your graph to make it more presentable (consider axes, colors, themes)
```{r}
ggplot(starwars, aes(eye_color)) + 
  geom_bar(stat= "count",
           width = 1,
            color = "#499607",
           fill = "#5FDD9D")+ 
  theme(
      plot.title = element_text(size = 17, face = "bold", color = "#8B1E3F", hjust = 0.5),
      axis.title.x = element_text(size = 13, face = "italic", color = "#1F2421"),
      axis.title.y = element_text(size = 13, face = "italic", color = "#1F2421"),
      axis.text.x = element_text(size = 8, face ='bold' ,angle = 45, hjust = 1)) +
  ggtitle("Star Wars Eye Color Frequency") + 
  xlab("Color") + 
  ylab("Frequency")
```

4. Create a graph of character's height versus mass.
```{r}
ggplot(starwars, aes(height,mass))+
  geom_point()
```

5. Customize this graph to make it more presentable.
```{r}
ggplot(starwars, aes(x = height, y = mass)) +
  geom_point(color = "#88498F", size = 2, alpha = 0.7) +  
  theme_minimal() +  
  theme(
      plot.title = element_text(size = 16, face = "bold", color = "#8B1E3F", hjust = 0.5),
      axis.title.x = element_text(size = 13, face = "italic", color = "#1F2421"),
      axis.title.y = element_text(size = 13, face = "italic", color = "#1F2421"),
      axis.text.x = element_text(size = 8, face ='bold' ,angle = 0, hjust = .5)
  ) +
  scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, 25)) +  # help deal with funky data points. Unavoidable to omit? 29 rows 
  scale_x_continuous(limits = c(50, 250), breaks = seq(50, 250, 25))  +
  xlab("Height (cm)") +
  ylab("Mass (kg)") +
  ggtitle("Height & Mass of Star Wars Characters")
```

6. Create another graph of character's height and mass. Include sex as a third variable.
```{r}
ggplot(starwars, aes(x = height, y = mass, color = sex)) +
  geom_point(size = 3, alpha = 0.7) + 
  theme_void()  # void theme probably makes it worse than it has to be... 
```

7. Customize this graph to make it more presentable. 
```{r}
# remove missing to help plot 
clean_starwars <- starwars %>%
  filter(!is.na(sex))

ggplot(clean_starwars, aes(x = height, y = mass, color = sex)) +
  geom_point(size = 3, alpha = 0.45) + 
  theme_minimal() + # why won't apa work? Updated ggplot... 
  theme( 
      plot.title = element_text(size = 16, face = "bold", color = "#CC8B86", hjust = 0.5),
      plot.subtitle = element_text(size = 12, face = "italic", color = "black", hjust = 0.5),
      axis.title.x = element_text(size = 11, face = "italic", color = "#1F2421"),
      axis.title.y = element_text(size = 11, face = "italic", color = "#1F2421"),
      axis.text.x = element_text(size = 8, face ='bold' ,angle = 0, hjust = .5)) +
  scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, 25)) +  # help deal with funky data points. Unavoidable to omit? 29 rows 
  scale_x_continuous(limits = c(50, 250), breaks = seq(50, 250, 25))  +
  scale_color_manual(
    values = c("#ef2d56", "#29335C", "#8cd867", "#ed7d3a")) +
  labs(
    x = "Height (cm)",
    y = "Mass (kg)",
    title = "Star Wars Character Height and Mass",
    subtitle = "Sorted by Sex",
    color = "Character Sex" # why color for legend title? not logical 
  )
```

