---
title: "Statistics in R Assignment"
author: "Amanda Mae Woodward"
date: "2024-11-05"
output: html_document
---

1. Open the diamonds dataset from ggplot2.  
```{r}
library(ggplot2)
data(diamonds)
head(diamonds)
```

2. Run a correlation between the x and y variables. What do you notice? 
```{r}
cor.test(diamonds$x,diamonds$y)
# These two variables are highly correlated, 0.97. There is almost a perfect positive linear relationship between the x and y variables from the dataset 
```

3. What is the correlation between price and carat?
```{r}
cor.test(diamonds$carat, diamonds$price)
# These two variables also have a strong positive linear relationship, 0.92. 
```

4. Create a linear model predicting price from table. What is the coefficient for table? 
```{r}
priceTabMod <- lm(price ~ table, data = diamonds)
summary(priceTabMod)
# The coefficient for the table variable is 226.984. 
```

5. Run a regression and an Anova predicting price from color. What do you notice about the two types of output? 
```{r}
# create linear model 
priceColorMod <- lm(price ~ color, data = diamonds)
summary(priceColorMod)

# create anova model 
priceColorANOVA <- aov(price ~ color, data = diamonds)
summary(priceColorANOVA)

# The output of these functions will be the same since they are mathematically identical 
```

6. Run a t test to determine if "D" color diamonds are more expensive than "E" color diamonds. 
```{r}
# First, going to need to subset the data since the color data is a leveled factor 
d_dia <- subset(diamonds, color == "D")
e_dia <- subset(diamonds, color == "E")

t.test(d_dia$price, e_dia$price, alternative = "greater")
# summary()

# Based on the results of the statistical test, 'D' color diamonds are typically more expensive than 'E' color diamonds. 
```

7. Open the warpbreaks dataset. What test would you use to determine the correlation between wool type and breaks?
```{r}
data(warpbreaks)
head(warpbreaks)
# To find the correlation between these two variables I would use a Pearson Correlation. However, I think I would need to change the wool type either to a numeric or a factor so the function would be able to handle the data class 
```

8. What is the correlation between wool type and breaks? 
```{r}
cor.test(as.numeric(warpbreaks$wool),warpbreaks$breaks)
# The correlation is negative, and -0.22. 
```

9. A student has a 35 question, multiple choice test. Each question has 4 choices. If they answered 17 questions correctly, did their responses differ from chance?  
```{r}
studentAns <- binom.test(17, 35, p = 0.25, alternative = "greater")
studentAns
# The response of the student did not differ from chance, since our p-value exceeds 0.05.  
```

