---
title: "psy4802_finalProjectAnalysis"
author: "Brady M. Chisholm"
date: "2024-11-25"
output: html_document
---

## Overview

In this pupil dilation data, measurements are one subjects average under specific combinations of conditions (4 SNR conditions, 5 quintiles), averaged over the 7-9 seconds time stamp in a trial. 

This results in 20 measurements per subject. For example, the first subject in this data set is P02, occupying the first 20 rows, followed by P03, etc. 

This analysis serves as an initial exploration. Ideally, the data should include trial-by-trial measurements (not averages), which would significantly increase temporal resolution and the quality of our statistics. 

Future analyses will extend from this project using raw data instead of the averages, and depending on results, a different model too.

### Load data, prep

```{r prep}
library(dplyr)
library(ggplot2)
library(mgcv) 
library(lme4) 
library(boot)

# Load data
url <- "https://raw.githubusercontent.com/bchisholm08/psy4802_fa24/5ffb123f8c664394d684cdb64290c003aebb085f/Final%20Project/Final%20Project%20Full%20Documentation/earlyVersions/quintileData.csv"
pupilDat <- read.csv(url)
```

### Data Cleaning
```{r cleaning}
# clean data, prepare for stats 
pupilDat$SNR <- as.numeric(as.character(pupilDat$SNR))
pupilDat$Quintile <- as.numeric(as.character(pupilDat$Quintile))
pupilDat$SubjID <- as.factor(pupilDat$SubjID)
```

### Summary Statistics
```{r summary stats}
# summary stats for SNR and Quintile. Check for data issues
summary_stats <- pupilDat %>%
  group_by(SNR, Quintile) %>%
  summarise(
    Mean_Dilation = mean(AvgPupilDilation, na.rm = TRUE),
    Med_Dilation = median(AvgPupilDilation, na.rm = TRUE),
    SD_Dilation = sd(AvgPupilDilation, na.rm = TRUE))
# print summary stats table 
print(as.data.frame(summary_stats))
```

### Check data distribution
```{r data dist}
# Plot histogram w/ density overlay 
ggplot(pupilDat, aes(x = AvgPupilDilation)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "#A0B9C6") +
  geom_density(alpha = 0.7, fill = "#802392") +
  labs(
    title = "Avg Pupil Dilation",
    subtitle = "Density plot in purple",
    x = "Avg Pupil Dilation (a.u.)",
    y = "Frequency"
  )
```

```{r originalDat corr matrix}
# Corr matrix for og data
originalDataCorr <- stats::cor(pupilDat %>% select_if(is.numeric), use = "complete.obs")
print(originalDataCorr)
```

### GAM Models

note; k = knots. Also called lambda. K = number of levels in variable. It tells the GAM function 
 
num of knots dictates flexibility of the smoothing function. Too few knots can underfit; too many can overfit.

num knots based on the complexity of the interaction and stats efficiency/computing power
```{r GAM mod1}
# for mod 1, regular GAM with no interactions
mod1_gam <- gam(AvgPupilDilation ~ s(SNR, k = 4) + s(Quintile, k = 5), data = pupilDat)
summary(mod1_gam)
plot(mod1_gam, pages = 1) # groupme
```

```{r} 
gam.check(mod1_gam) # model diagnostic plots 
```

### Mod2
```{r GAM mod2}
# mod 2; use tensor interaction of SNR & quintile 
mod2_gam <- gam(AvgPupilDilation ~ te(SNR, Quintile, k = 4), data = pupilDat)
summary(mod2_gam)
plot(mod2_gam, pages = 1)

gam.check(mod2_gam) 
```

### Mod3
```{r GAM mod3}
# mod 3, both interactions and random effects 
mod3_gam <- gam(AvgPupilDilation ~ te(SNR, Quintile, k = 4) + s(SNR, k = 4) + s(Quintile, k = 5) + s(SubjID, bs = "re"), data = pupilDat)
summary(mod3_gam)
plot(mod3_gam, pages = 1)
gam.check(mod3_gam)
```

### Compare models 
```{r compare GAMs}
gamMods_comparison <- data.frame(
  Model = c("Mod_1", "Mod_2", "Mod_3"),
  AIC = c(AIC(mod1_gam), AIC(mod2_gam), AIC(mod3_gam)),
  BIC = c(BIC(mod1_gam), BIC(mod2_gam), BIC(mod3_gam))
)

knitr::kable(gamMods_comparison, digits = 3, caption = "AIC & BIC Comparison")
#FIXME EXPLAIN AIC + BIC 
```

# LME 
```{r LME mod}
# for fun, check LME model
mod4LMER <- lmer(AvgPupilDilation ~ SNR * Quintile + (1 | SubjID), data = pupilDat)
summary(mod4LMER)
```

### plot LME results
```{r lme plot}
# scatter pupil & quintile
ggplot(pupilDat, aes(x = Quintile, y = AvgPupilDilation)) +
  geom_smooth(method = "loess", color = "#6B8F71", se = TRUE) +
  labs(
    title = "Avg Pupil Dilation vs. Quintile",
    x = "Quintile",
    y = "Avg Pupil Dilation (a.u.)"
  ) +
  theme_minimal()
```

# Bootstrap data

One concern with the GAM is that there is not enough data. In the introduction we discussed how the data is organized, with each subject getting 20 rows for each combination of SNR and quintile. This gives us just 47 observations of each condition, which may not be enough for the GAM.

```{r bootstrap}
# bootstrap function
bootstrap_sample <- function(data, indices) {
  d <- data[indices, ]
  return(d)}
# bootstrap data & allow replacement
set.seed(4802)
bootstrapped_data <- do.call(rbind, 
                             replicate(50, pupilDat[sample(1:nrow(pupilDat), replace = TRUE), ], simplify = FALSE))
```
use do.call for function args

note: original data is 20 observations per subject. Bootstrapping assumes each subject is observed just once, so something is not right with the bootstrapped data as a result. 

Rerun models with Bootstrapped Data. Small sample size could be creating some issues for a GAM. 

### Bootstrapped cor matrix
```{r redo corr}
# compare correlations
print(cor(bootstrapped_data %>% select_if(is.numeric), use = "complete.obs"))
print(originalDataCorr)
```
Compare correlations in bootstrapped vs original data before proceeding to modeling

### bootstrap mod1; mod2; mod3
```{r GAM w/bootstrap}
# no interactions
mod1_boot <- gam(AvgPupilDilation ~ s(SNR, k = 4) + s(Quintile, k = 5), data = bootstrapped_data)
summary(mod1_boot)
plot(mod1_boot, pages = 1)

# bootstrap mod2 only interaction 
mod2_boot <- gam(AvgPupilDilation ~ te(SNR, Quintile, k = 4), data = bootstrapped_data)
summary(mod2_boot)
plot(mod2_boot, pages = 1)

# mod3 full bootstrap model 
mod3_boot <- gam(AvgPupilDilation ~ te(SNR, Quintile, k = 4) + s(SNR, k = 4) + s(Quintile, k = 5) + s(SubjID, bs = "re"), data = bootstrapped_data)
summary(mod3_boot) 
plot(mod3_boot, pages = 1)

#FIXME explain s() and te() functions, bs = "re"
```

```{r gpt test code}
bootMods <- data.frame(
  Model = c("Mod_1_Boot", "Mod_2_Boot", "Mod_3_Boot"),
  AIC = c(AIC(mod1_boot), AIC(mod2_boot), AIC(mod3_boot)),
  BIC = c(BIC(mod1_boot), BIC(mod2_boot), BIC(mod3_boot)),
  Equation = c(
    "AvgPupilDilation ~ s(SNR, k=4) + s(Quintile, k=5)",
    "AvgPupilDilation ~ te(SNR, Quintile, k=4)",
    "AvgPupilDilation ~ te(SNR, Quintile, k=4) + s(SNR, k=4) + s(Quintile, k=5) + s(SubjID, bs='re')"
  )
)
bootMods
```


#```{r model comp.}
# compare bootstrap models to real models
bootMODS <- data.frame(
  Model = c("Mod_1_Boot", "Mod_2_Boot", "Mod_3_Boot"),
  AIC = c(AIC(mod1_boot), AIC(mod2_boot), AIC(mod3_boot)),
  BIC = c(BIC(mod1_boot), BIC(mod2_boot), BIC(mod3_boot)),
    Equation = c(
"AvgPupilDilation ~ s(SNR, k=4) + s(Quintile, k=5)",
"AvgPupilDilation ~ te(SNR, Quintile, k=4)",
"AvgPupilDilation ~ te(SNR, Quintile, k=4) + s(SNR, k=4) + s(Quintile, k=5) + s(SubjID, bs='re')", 
"AvgPupilDilation ~ s(SNR, k=4) + s(Quintile, k=5)",
"AvgPupilDilation ~ te(SNR, Quintile, k=4)",
"AvgPupilDilation ~ te(SNR, Quintile, k=4) + s(SNR, k=4) + s(Quintile, k=5) + s(SubjID, bs='re')"))

combined_comparison <- rbind(
  cbind(gamMods_comparison, Source = "Original"),
  cbind(bootMODS, Source = "Bootstrapped")
)

knitr::kable(combined_comparison, digits = 3, caption = "Model Comparison: Original vs Bootstrapped Data")
#```

Models were rerun on the bootstrapped dataset to increase sample size and improve stability. The comparison of AIC and BIC metrics provides insights into whether bootstrapping improves model performance. Interpretations should consider both fit and overfitting risks.

### Exploratory, plot LME model with bootstrapped data 
```{r lme bootstrap}
LMER_boot <- lmer(AvgPupilDilation ~ SNR * Quintile + (1 | SubjID), data = pupilDat)
summary(LMER_boot)

# scatter pupil & quintile
ggplot(bootstrapped_data, aes(x = Quintile, y = AvgPupilDilation)) +
  geom_smooth(method = "loess", color = "#B57BA6", se = TRUE) +
  labs(
    title = "Avg Pupil Dilation vs. Quintile",
    x = "Quintile",
    y = "Avg Pupil Dilation (a.u.)"
  ) +
  theme_minimal()
```

